Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-04 TOMS_Paper/assembly.tex
Add introduction

@@

\subsection{Code Generation for Quadrature Representation}
Rapid implementation of high performance, robust, and portable code evaluating element matrices using quadrature can be achieved through automated code generation. This has been successfully proved in the context of the popular FEniCS project~\cite{Fenics}. The FEniCS Form Compiler (FFC) accepts as input the variational form of a partial differential equation and generates C++ code implementing local assembly routines. The variational form is expressed at high-level by means of the domain-specific Unified Form Language (UFL). Local assembly code must be high performance: as the complexity of a form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen functions, the resulting kernels evaluating element matrices become more computationally expensive, which impacts significantly the run-time of the overall computation.

Achieving high performance is non-trivial due to the complexity of the mathematical expressions involved in the numerical integration and because of the small sizes of loops and accessed arrays. In~\cite{quadrature1},~\cite{Kirby-FEM-opt} and~\cite{Francis}, it is shown how automated code generation allows the introduction of powerful optimizations, which a user cannot be expected to write ``by hand'', as well as the exploration of non-standard integration techniques, based, for instance, on symbolic manipulation. In~\cite{Luporini} we made one step forward by showing that different problems require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations, which are not supported by available compilers, are essential to maximize instruction-level parallelism and register locality.

\subsection{Summary of Low-level Optimizations for Quadrature Representation}
\label{sec:summary-opts}
To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work on automated code transformations for quadrature representation.

@@

The work in~\cite{Luporini} resulted in the development of COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, a compiler for the optimization of local assembly kernels relying on quadrature representation.

@@

\section{A Compiler for Optimizing Quadrature-based Integration}
In order to generate high performance code, mathematical expressions evaluating the element tensor must be optimized with regards to several interrelated aspects: 1) minimization of floating point operations, 2) instruction-level parallelism, and 3) data locality. In this paper, we tackle these three points building on our previous work (\cite{Luporini}).

@@

...TODO...: to what extent applying rewrite rules, is problem specific...
% blablabla molto complesse, nestings etc expose etc allora rewrite rules

>>>>

% Quadrature. Tensor. Quadrature <, >, = Tensor
%code evaluating element matrices using quadrature

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become incredibly popular over the last years. In these frameworks, the weak variational form of a given problem is expressed at high-level by means of a domain-specific language. Such a mathematical specification is suitably manipulated and then passed as input to a form compiler, whose goal is to generate a representation of local assembly operations. These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution. Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more expensive, covering a significant fraction of overall computation run-time.

Achieving high performance implementations is, however, non-trivial. The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process. Also, traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions. This has led to development of a number of higher-level approaches to optimize local assembly kernels.  In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''. \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of the problem. In~\cite{Luporini}, we have made one step forward by showing that different problems, on different platforms, require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality. The problem of optimizing local assembly routines has been tackled recently also for GPUs, for instance in~\cite{petsc-integration-gpu}.

Our research has resulted in the development of a compiler, COFFEE\footnote{COFFEE stands for COmpiler For FinitE Element local assembly.}, fully integrated with the Firedrake framework. While clearly separating the mathematical domain, which remains captured in the higher-level form compiler, from the optimization process, COFFEE also aims to be platform-agnostic. The code transformations occur on an intermediate representation of the assembly code, which is ultimately translated into platform-specific code. Domain knowledge is exploited in two ways: for simplifying the implementation of a broad range of code transformations, and, obviously, to make them extremely effective. Domain knowledge is conveyed to COFFEE from the higher level (the form compiler in the case of Firedrake, although any user-provided code would be acceptable) through suitable annotations attached to the input. For example, when the input is in the form of an abstract syntax tree produced by the form compiler, specific nodes are decorated so as to drive the optimization process. Although COFFEE has been thought of as a multi-platform optimizing compiler, our performance evaluation so far has been restricted to standard CPU platforms only. We emphasize once more, however, that the transformations applicable by both the Expression Rewriter (Section~\ref{sec:expr-rewriting}) and the Code Specializer (Section~\ref{sec:code-spec}) would work on generic accelerators as well.

In this paper, we build on the work presented in~\cite{Luporini} and present a novel structured approach to the optimization of automatically-generated finite element integration routines. We argue that peak performance can be achieved only by passing through a two-step optimization procedure: 1) expression rewriting, to minimize floating point operations, 2) and code specialization, to obtain, for instance, effective register utilization and SIMD vectorization. The code transformations introduced in~\cite{Luporini} are reused: as explained in Section~\ref{sec:summary-opts}, padding and data alignment, expression splitting, and vector-register tiling become sub-steps of code specialization; on the other hand, generalized loop-invariant code motion is a step of the expression rewriting process. More importantly, we complement and generalize our previous work with the following contributions.

Expression rewriting is based on a formal set of rewrite rules. Our first contribution consists of a framework that aggressively exploits associativity, distributivity, and commutativity of arithmetic operators to expose ``hidden'' loop-invariant sub-expressions. Secondly, we show how to make use of domain knowledge to avoid computation over zero-valued regions in vector-valued basis functions arrays, while preserving code vectorizability. These transformations will allow outperforming the results obtained in~\cite{Luporini} as well as those achievable by using FEniCS' built-in optimizations, presented in~\cite{quadrature1}.

At code specialization time, transformations are applied to maximize the exploitation of the underlying platform's resources, e.g. SIMD lanes. On top of the work in~\cite{Luporini}, we provide a number of contributions. Firstly, we show the benefit of vector-expansion to achieve SIMD vectorization of otherwise scalar code. This is particularly useful in complex forms, like those based on hyperelasticity. Secondly, we answer an open problem in~\cite{Luporini} by providing an algorithm that automatically transforms an element matrix evaluation into a sequence of calls to BLAS' dense matrix multiplies. BLAS routines are known to perform far from peak performance when the involved arrays are small, which is almost always the case of low-order finite element methods. However, we will show that in corner, yet important cases, especially in forms characterized by pre-multiplying functions and relatively high-order function spaces, a BLAS-based execution strategy can be a successful. Finally, we introduce a model-driven, dynamic autotuner that automatically and transparently compose the set of code transformations that are likely to maximize the performance of a given problem. The main challenge with the autotuner is to maintain, for any possible problem, the search space reasonably small, although comprising the most effective code variants, so that the overhead, which impacts the run-time, is negligible.

Expression rewriting and code specialization have been implemented in COFFEE and are fully operating. Therefore, to testify the goodness of our approach, we provide an extensive and unprecedented performance evaluation in a number of forms of increasing complexity, including problems based on hyperelasticity operators. We characterize our problems by varying polynomial order of the employed function spaces and number of pre-multiplying functions. To clearly distinguish the improvement achieved by this work, we will compare four sets of code variants, for each problem instance: 1) unoptimized code, i.e. a local assembly routine as returned from the form compiler; 2) code optimized by FEniCS, i.e. the work in~\cite{quadrature1}; 3) code optimized as described in~\cite{Luporini}; code optimized by expression rewriting and code specialization as described in this paper. Notable performance improvements of 4) over 1), 2) and 3) are reported and detailed.

%The contributions of this paper can be summarized as follows:
%\begin{enumerate}
%\item 
%\item extensive and unprecedented performance evaluation in a wide range of forms of increasing complexity
%\end{enumerate}

%Extensive performance evaluation...

@@

\subsection{COFFEE: a Compiler for Optimizing Quadrature-based Finite Element Integration}
In order to generate high performance code, mathematical expressions evaluating the element tensor must be optimized with regards to several interrelated aspects: 1) minimization of floating point operations, 2) instruction-level parallelism, and 3) data locality. In this paper, we tackle these three points building on our previous work (\cite{Luporini}).

\subsection{Summary of Optimizations for Automated Quadrature Representation}
\label{sec:summary-opts}
...TODO... : clearly state, once more, we operate in the regime of automated code generation

To neatly distinguish the contributions of this paper from those in~\cite{Luporini}, in this section we summarize the results of our previous work on automated code transformations for quadrature representation.

@@

The work in~\cite{Luporini} resulted in the development of COFFEE, a compiler for the optimization of local assembly kernels relying on quadrature representation.

@@

@@

...TODO...: to what extent applying rewrite rules, is problem specific...

...TODO...: things get interesting if the kernel operates on a batch of elements...can save a LOT of flops
% blablabla molto complesse, nestings etc expose etc allora rewrite rules
